---
title: Statistical Analysis of ENEM 2019 Grades in the state of Alagoas
author: Isadora Rodrigues
site: "bookdown::bookdown_site"
output:
  prettydoc::html_pretty:
    theme: architect
    highlight: github
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# **Importing the libraries**

```{r message=FALSE}
library(readxl)
library(ggplot2)
library(dbplyr)
library(tidyverse)
library(rstatix)
library(Hmisc)
library(ggpubr)
library(corrplot)
library(ggprism)
```

# **Importing the data**

```{r}
df = read_xlsx('data/enem.xlsx')
```

```{r}
cat('Mean = ',mean(df$NOTA_ENEN))
```

# Grades distribution

```{r}
hist(df$NOTA_ENEN, col = "lightblue", main = "", xlab = "Frequency", 
     ylab = "Grades", breaks = 30)
abline(v=mean(df$NOTA_ENEN), col='red',lwd=2)
legend(x='topright',legend=paste('Mean = ',signif(mean(df$NOTA_ENEN))), fill='red')
```

The graph above shows a high frequency of grades between 400 and 500, with an average of 509,144.

# Grades distribution by gender

```{r}
grades_cut = cut(df$NOTA_ENEN, breaks = quantile(df$NOTA_ENEN),
                include.lowest = TRUE)

quartiles_grade_gender = table(df$TP_SEXO, grades_cut)

barplot(quartiles_grade_gender, beside = TRUE, xlab = "Grade by quartiles", 
        ylab = "Frequency", col = c("lightblue", 5))

legend(x = "topright", legend = c("Female", "Male"),
       fill = c("lightblue", 5), bty = "n")
```

In the distribution of scores by quartiles in female and male candidates, we can see that males have the lowest scores.

Another observation is that there is an adverse relationship in both sexes, while in females, the higher the grade, the lower the frequency falls, in males, the higher the grade quartile, the higher the frequency.

# Statistical difference tests between men's and women's grades for the Writing grade

*H0*: there is no statistical difference;

*H1*: there is statistical difference.

## Separating variables

```{r}
women_grades = subset(df$NOTA_ENEN,df$TP_SEXO=='Feminino')
men_grades = subset(df$NOTA_ENEN,df$TP_SEXO=='Masculino')
```

## Variation test

```{r}
var.test(men_grades,women_grades)
```

As p-value is greater than 0.05, then hypothesis H0 is not rejected, that is, there is evidence at 5% of significance that the variances are equal. In this way they can be combined to calculate an estimate of the standard deviation.

## T test

```{r}
t_test=t.test(men_grades,women_grades, conf.level = 0.05)
t_test
```

As p-value is greater than 0.05, then hypothesis H0 is not rejected, that is, there is evidence at 5% of significance that the means are not different.

**Thus, there is no statistical difference between the essay grades between men and women.**

## Graph illustrating the analysis of variance of grades between genders

```{r}
df_p_val <- data.frame(
  group1 = "Female",
  group2 = "Male",
  label = signif(t_test$p.value, digits=3),
  y.position = 1000
)

ggboxplot(df, x = "TP_SEXO", y = "NU_NOTA_REDACAO",
          xlab = 'Gender', ylab = 'Writing Grades') +
  add_pvalue(df_p_val)
```


# Classification of notes in Mesoregions

```{r}
meso = read_xlsx('data/mun_messoregiao.xlsx')
mean_mun=aggregate(NOTA_ENEN~NO_MUNICIPIO_RESIDENCIA,data=df,FUN=mean)
mean_mun_meso=merge(mean_mun,meso,by.x='NO_MUNICIPIO_RESIDENCIA',by.y='mun')
```

```{r echo=FALSE}
knitr::kable(mean_mun_meso[1:10,], col.names = c('Counties','Grade','Mesoregions'))
```

## Distribution of grades by mesoregion

```{r}
grade_cut_m = cut(mean_mun_meso$NOTA_ENEN, breaks = quantile(mean_mun_meso$NOTA_ENEN),
                 include.lowest = TRUE)

quartiles_grade = table(mean_mun_meso$mesorregiao, grade_cut_m)

barplot(quartiles_grade, beside = TRUE, col = c("lightblue", 5, 'blue'), 
        ylab = "Frequency", xlab = "Grades by quatiles")

legend(x = "topleft", 
       legend = c("Wild", "Sertão", "East"), 
       fill = c("lightblue", 5, 'blue'), xpd = TRUE)
```

In the distribution of scores by mesoregion, we can see that the sertão stratum has the highest score in all quartiles and that the eastern region has a higher frequency of low scores.

# Statistical difference tests between the scores of the mesoregions

## Analysis of variance between the means of the mesoregions

**For this analysis, the ANOVA test was initially used**

*H0*: the means of the groups are the same

*H1*: at least one mean of the groups is different from the others

```{r warning=FALSE, results='hide', message=FALSE}
res.aov=mean_mun_meso%>%
  anova_test(NOTA_ENEN ~ mesorregiao)
```

```{r echo=FALSE}
knitr::kable(res.aov)
```

As p \> 0.05, the null hypothesis is not rejected, that is, there are no significant differences between the groups (p = 0.264).

**To analyze the mesoregions in pairs, the pairwise t test was used.**

*H0*: the averages of the pair are the same

*H1*: the averages of the pair are different

```{r}
pwc <- mean_mun_meso %>%
  pairwise_t_test(NOTA_ENEN ~ mesorregiao, p.adjust.method = "bonferroni")
```

```{r echo=FALSE}
knitr::kable(pwc)
```

It can be seen that all p.adj were greater than 0.05, that is, **there is no statistical difference between the averages of the mesoregions.**

## Graph illustrating the analysis of variance

```{r}
pwc <- pwc %>% add_xy_position(x = "mesorregiao")
ggboxplot(mean_mun_meso, x = "mesorregiao", y = "NOTA_ENEN",
          xlab = 'Mesoregion', ylab = 'Grades') +
  stat_pvalue_manual(pwc, label = "p.adj", tip.length = 0, step.increase = 0.1,) +
  labs(
    subtitle = get_test_label(res.aov, detailed = TRUE),
    caption = get_pwc_label(pwc)
    )
```

# Correlation matrix between the scores of the 5 ENEM subjects

```{r}
grande_cn = aggregate(NU_NOTA_CN~NO_MUNICIPIO_RESIDENCIA,data=df,FUN=mean)
grades = merge(grande_cn,aggregate(NU_NOTA_CH~NO_MUNICIPIO_RESIDENCIA,data=df,
                                FUN=mean),by='NO_MUNICIPIO_RESIDENCIA')
grades = merge(grades,aggregate(NU_NOTA_LC~NO_MUNICIPIO_RESIDENCIA,data=df,
                              FUN=mean),by='NO_MUNICIPIO_RESIDENCIA')
grades = merge(grades,aggregate(NU_NOTA_MT~NO_MUNICIPIO_RESIDENCIA,data=df,
                              FUN=mean),by='NO_MUNICIPIO_RESIDENCIA')
grades = merge(grades,aggregate(NU_NOTA_REDACAO~NO_MUNICIPIO_RESIDENCIA,
                                data=df,FUN=mean),by='NO_MUNICIPIO_RESIDENCIA')


rownames(grades) = grades$NO_MUNICIPIO_RESIDENCIA
grades$NO_MUNICIPIO_RESIDENCIA = NULL

corrplot(cor(grades), method = 'number')
```

Analyzing the correlation matrix, it is possible to observe that the subjects that presented Pearson's coefficient are Natural Sciences and Mathematics.

From this, we can perform a linear regression to identify the prediction equation.

## Regression equation

```{r}
cn_mt=select(grades,NU_NOTA_CN,NU_NOTA_MT)
plot(cn_mt,main='Notes Natural Sciences and Mathematics')

regression = lm(cn_mt$NU_NOTA_MT~cn_mt$NU_NOTA_CN)

abline(regression, col='red')
grid()
summary(regression)
```

Equation defined as: 1,36\*x-126,21

## Hypothesis test to validate the correlation

*H0*: there is normality in the data

*H1*: there is no normality in the data.

```{r}
shapiro.test(regression$residuals)
```

The null hypothesis was not rejected and there is evidence that the residuals are **normally distributed.**
